---
title: "Learning GPs"
format: html
---

### Simulate data from a GP! 

Everything is well explained in Mike's case study, but the following code will give you some ideas how to generate functional behaviors from a GP in practice.

#### The covariance function

Here we focus on the exponentiated quadratic covariance function, that can be written in R as:

```{r}
set.seed(12345)

k <- function(delta_x, gamma, rho){
  return(gamma^2*exp(-1/2*(delta_x/rho)^2))
}
```

Which allows to plot a covariogram like this (see Mike's case study):

```{r}

delta_x <- seq(0, 10, 0.5)
plot(x = delta_x,
     y = k(delta_x, gamma = 1, rho = 4), type = 'l',
     bty = 'n', col = "#487575",
     xlab = expression(Delta*'x'), ylab = 'Covariance')

```

#### From covariance to sampling

With a finite number of covariate values for which we want to make predictions, the finite-dimensional projection of a GP essentially become a multinormal distribution. The covariance matrix of this distribution is determined by the covariance function.

For example, let's say we have a grid 10 points evenly distributed fromn 1 to 10. We can compute a distance matrix as follows:

```{r}

grid <- data.frame(x1 = seq(1,10,1), x2 = seq(1,10,1))
dmat <- as.matrix(dist(grid, diag = T, upper = T))

print(round(dmat, 2))

```

It becomes then straightforward to apply the covariance function to this matrix, for the grid of points we consider:


```{r}

cov <- apply(dmat, 1:2, k, gamma = 1, rho = 7)

print(round(cov, 2))

```

Using this covariance matrix, we can then simulate some GP functional behaviors along this grid:

```{r}

f <- MASS::mvrnorm(20, rep(0,10), cov)

plot(x = seq(1,10,1), y = f[1,], 
     type = 'l', bty = 'n', col = "#B97C7C",
     ylim = c(-3,3), xlim = c(1,10),
     xlab = 'x', ylab = 'f(x)')
for(i in 2:20){
  lines(x = seq(1,10,1), y = f[i,], col = "#B97C7C")
}


```

We can repeat the process with different parameters for the covariance function:

```{r}

cov <- apply(dmat, 1:2, k, gamma = 2, rho = 1.5)
f <- MASS::mvrnorm(20, rep(0,10), cov)

plot(x = seq(1,10,1), y = f[1,], 
     type = 'l', bty = 'n', col = "#B97C7C",
     ylim = c(-6,6), xlim = c(1,10),
     xlab = 'x', ylab = 'f(x)')
for(i in 2:20){
  lines(x = seq(1,10,1), y = f[i,], col = "#B97C7C")
}


```

This looks a bit discontinuous as we sampled only a grid of 10 points.\
If we increase the resolution of the grid, we obtain something smoother.

```{r}
n <- 1000

grid <- data.frame(x1 = seq(1,10, length.out = n), x2 = seq(1,10, length.out = n))
dmat <- as.matrix(dist(grid, diag = T, upper = T))

cov <- apply(dmat, 1:2, k, gamma = 2, rho = 1.5)
f <- MASS::mvrnorm(20, rep(0,n), cov)

plot(x = seq(1,10, length.out = n), y = f[1,], 
     type = 'l', bty = 'n', col = "#B97C7C",
     ylim = c(-6,6), xlim = c(1,10),
     xlab = 'x', ylab = 'f(x)')
for(i in 2:20){
  lines(x = seq(1,10, length.out = n), y = f[i,], col = "#B97C7C")
}


```

#### Simulate observations that arise from a GP

What if we have a normal observational model with a variance $\sigma^2$? Let's say that the functional relationship between our observations is the following GP:

```{r}
n <- 1000

grid <- data.frame(x1 = seq(1,10, length.out = n), x2 = seq(1,10, length.out = n))
dmat <- as.matrix(dist(grid, diag = T, upper = T))

cov <- apply(dmat, 1:2, k, gamma = 1.7, rho = 2.7)
f <- MASS::mvrnorm(1, rep(0,n), cov)

plot(x = seq(1,10, length.out = n), y = f, 
     type = 'l', bty = 'n', col = "#B97C7C",
     ylim = c(-6,6), xlim = c(1,10),
     xlab = 'x', ylab = 'f(x)')

```

We can then simulate observations with a measurement error $\sigma$:

```{r}

# for 13 observations
idxs <- sample(1:n, 13)
x <- seq(1,10, length.out = n)[idxs] 
fx <- f[idxs]

y <- rnorm(13, fx, sd = 1)

plot(x = seq(1,10, length.out = n), y = f, 
     type = 'l', bty = 'n', col = "#B97C7C",
     ylim = c(-6,6), xlim = c(1,10),
     xlab = 'x', ylab = 'f(x)')
points(x = x, y = y,
       pch = 20, col = "#8F2727")

```